### 2012-10-17 rnaseq_paper.tex

We used the Velvet\cite{Zerbino:2008vu} and Oases\cite{Schulz:2012je}
assemblers to construct transcript fragments from four entire Illumina
GAII mRNAseq data sets sequenced from chicken spleen (see Methods and
Materials).  In the assembly, we used multiple distinct k-mer values
for Velvet because different isoforms are recovered for different $k$
\cite{Schulz:2012je}.  We chose k-mers between 21 and 31 for this
study, and recovered between 340,000 and 370,000 unique sequences from
each data set (see Table~\ref{unique_sequences}, and Materials and
Methods).  These unique sequences represented an unknown number of
true genes, due to fragmentation from low coverage and incomplete
assembly.

We next used Tophat to align mRNAseq reads to the genome and partition
reads by chromosome; we then assembled the partitioned reads with
Velvet and Oases using the same range of parameters as the global
assembly, above.  While the local assemblies were considerably more
computationally efficient, they lacked several thousand unique regions
that were present in the global assembly
(Table~\ref{unique_sequences_matched_mouse}); this is probably due to
the incomplete nature of the current chicken genome (cite).
Interestingly, over a hundred regions were present {\em only} in the
local assemblies, suggesting that the local assemblies might be
recovering additional exons.  Significant numbers of unique regions
($>300$bp) from both global and local assemblies showed homology to
the mouse genome, indicating that at least some of these unique sequences
represented real sequence content.  Figure~\ref{global_vs_local} shows
an example of different isoforms detected by the two assembly methods.

@@

@@

To construct a more comprehensive gene set containing all of the
assembled contigs, we tried using Oases-M to merge the assemblies from
multiple $k$ values \cite{Schulz:2012je}.  While it has been
demonstrated that merged transcripts from multiple-k assemblies
contain more isoforms than those from any single $k$, the sensitivity
of Oases-M for recovering splice variants has not been fully
evaluated.  We merged transcripts from our global assembly, above,
with Oases-M using a k-mer size of 27, and then compared them with the
unmerged transcripts.  We then cross-validated using publicly
available ESTs for an independent data set.  The results show that
Oases-M and the unmerged assembly share about 95\% of the predicted,
splice junctions, with about 5\% disjoint.  Of these 5\%,
approximately 6\% of the Oases-M-specific splice junctions are
independently supported by ESTs, while 19\% of the unmerged splice
junctions are supported by ESTs.  This suggests that Oases-M probably
discards a number of real splice variants, although the unmerged
assembly is also probably missing some found by Oases-M.

@@

We developed an exon graph approach to construct gene models from
alignments of transcripts against the genome.  Our approach, called
Gimme, uses exon graphs to merge transcripts and gene models based on
overlapping exons (see Materials and Methods).  We used Gimme to
obtain 20,439 gene models containing 26,035 isoforms from our global
assembly; 21,963 gene models containing 25,859 from our local
assembly; and 20,922 gene models containing 27,732 isoforms from the
merged global and local assemblies (Table~\ref{genes_transcripts}).
The number of genes is slightly decreased after merging transcripts
from global and local assembly together. 
This indicates that some transcripts from both methods were merged together.
Figure~\ref{model_comparisons} shows gene models built by our method with
other existing gene models.

@@

The results show that our pipeline predicts many gene models and
isoforms after assembly.  We validated these gene models in several
different ways.

@@

We used bowtie to map single-end reads from the source datasets to the
transcripts.  More than 80\% of the original reads could be aligned to
the transcripts, demonstrating that we did not lose a significant
amount of information during the merge process
(Table~\ref{single-end_map}).  More importantly, we also mapped
paired-end reads from technical replicates to the same gene models,
and found that more than 85\% of the paired-ends mapped concordantly
the gene models (Table~\ref{paired-end_map}).  Most of the reads that
did not map were either highly erroneous or contained low-complexity
artifactual sequence that probably came from the reverse transcription
procedure (Cite).
% @CTB do we have something we can cite for typical % mapping?  Maybe Trinity
% paper?  How about a citation that suggests that there's a lot of crap
% in mRNAseq?
Thus the merged gene models produced by Gimme represent the significant
majority of the assemblable data.

@@

To validate the splice junctions reconstructed by the Gimme pipeline,
we used Bowtie to map mRNAseq reads to the transcript sequences
derived from gene models Bowtie\cite{Langmead:2009fv}.  Because the
Velvet/Oases assembly pipeline does not make use of the reference
genome, reads that map across a splice junction constitute independent
verification of a splice junction's presence in a transcript.

Of 104,827 splice junctions from the gene models, 248 (0.23\%)
junctions have no spliced reads and only 484 (0.46\%) junctions
have fewer than 4 spliced reads (Fig.~\ref{cdf_single_splice}).
More than 99\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
truly present within the data set.  Note that lower coverage splice junctions
may still be correct, but may be from low-expressed isoforms or 
regions of the gene that are low coverage due to sample preparation issues.

@@

Of the 104,827 splice junctions in our gene models, 82,908 (80.31\%)
are supported by ESTs or mRNAs from Genbank.  This is especially
surprising since our mRNAseq data is from spleen, and few of the
publicly available ESTs or mRNAs are from spleen.  Note that this
cross-validation suggests that the 19,981 novel splice junctions {\em not}
seen in publicly available ESTs and mRNAs are also likely to be real
splice junctions from spleen.
% @CTB supported by ESTs OR mRNAs, or ESTs AND mRNAs?  Surely the former?

@@

The ENSEMBL pipeline does not, however, include a large number of
splice junctions from ESTs (99,755) or mRNAs (13,641).  Cufflinks and
Gimme each recover about 10\% of these, with more than 2/3 of these
recovered by both.  This indicates that both Gimme and Cufflinks
are equally adept at recovering novel splice junctions.
% @CTB note this suggests that Cufflinks improved recovery for ENSEMBL
% gene models is false...
% @CTB what fraction of Cufflinks splice junctions are high coverage (4+)?

@@

As shown above, Cufflinks and the assembly method detect a number of
distinct but equally valid splice junctions, which suggests that we
could obtain greater sensitivity by merging both sets of predictions.
We therefore used Gimme to merge mRNAs, the Cufflinks and assembly gene
models Table.~\ref{genes_transcripts}.  This resulted in an increased 
number of total genes and isoforms (e.g. see Fig.~\ref{mrna_cuff_gimme.pdf}).
The merged gene models recover 51\% and 66\% of splice junctions from ESTs
and ENSENBL respectively, which is 10-15\% greater than that from corresponding
unmerged gene models.
% @CTB what about splice junctions etc??

@@

@@

Most transcripts include a large part of untranslated regions (UTRs),
especially $3'$ UTR in our datasets.
These regions are challenging to predict correctly solely from computational
methods due to low degree of conservation [?].
RNA-Seq-derived gene models are useful for studying variations within UTR
regions, which may be involved in regulation of isoform expression\cite{}

@@

@@

@@

\subsubsection{isoforms and mouse}

@@

\section*{Tables}
\begin{table}[!ht]
\caption{
\bf{Unique sequences from global and local assembly}}
\begin{tabular}{ccccc}
\hline
Dataset & \multicolumn{2}{c}{Total Sequence} & \multicolumn{2}{c}{Unique Sequence}\\
 & Global & Local & Global & Local\\
\hline
Line 6 uninfected & 338,353 & 335,180 & 10,068 (2.97\%) & 1,258 (0.37\%)\\
Line 6 infected & 353,485 & 319,920 & 11,639 (3.29\%)& 2,270 (0.70\%)\\
Line 7 uninfected & 335,876 & 302,443 & 10,649 (3.17\%) & 1,570 (0.51\%)\\
Line 7 infected & 371,404 & 327,491 & 11,859 (3.19?\%)& 1,199 (0.36\%)\\
\hline
%table information
\end{tabular}
%\begin{flushleft}Table Caption
%\end{flushleft}
\label{unique_sequences}

@@

\caption{
\bf{Number of total and unique splice junctions}}
\begin{tabular}{cccccc}
\hline
Method& Total & Unique & Supported by ESTs \\ 
\hline
Oases-M assembly & 111,237 & 6,860 & 421 (6.1\%) \\
Unmerged assembly & 112,708 & 8,295 & 1,607 (19.4\%) \\
\hline
%table information
\end{tabular}
%\begin{flushleft}\footnotesize \textit{*Minumum set}
%\end{flushleft}
\label{Oases-M}

>>>>

We used the Velvet\cite{Zerbino:2008vu} and Oases\cite{Schulz:2012je}
assemblers to construct transcript fragments from four entire Illumina
GAII mRNAseq data sets sequenced from chicken spleen (see Methods and
Materials).  In the assembly, we used multiple distinct k-mer values
for Velvet to sensitively recover as many different isoforms as possible
\cite{Schulz:2012je}.  We chose k-mers between 21 and 31,
and recovered between 340,000 and 370,000 unique sequences from
each data set (see Table~\ref{unique_sequences}, and Materials and
Methods).  These unique sequences represented an unknown number of
true genes, due to fragmentation from low coverage and incomplete
assembly.

We next used Tophat to align mRNAseq reads to the genome and partition
reads by chromosome; we then assembled the partitioned reads with
Velvet and Oases using the same range of parameters as the global
assembly, above.  While the local assemblies were considerably more
computationally efficient, they lacked several thousand unique regions
that were present in the global assembly
(Table~\ref{unique_sequences_matched_mouse}); this is probably due to
the incomplete nature of the current chicken genome assembly, which is
lacking approximately 5\% of its true gene content (cite).
Interestingly, over a hundred regions were present {\em only} in the
local assemblies, suggesting that the local assemblies might be
recovering additional exons.  Significant numbers of unique regions
from both global and local assemblies showed homology to the mouse
genome, indicating that at least some of these unique sequences
represented real sequence content.  Figure~\ref{global_vs_local} shows
an example of different isoforms detected by the two assembly methods.

@@

% @CTB DISCSUSS the use of splice junctions for validity; transition
% from transcripts to splice junctions.

@@

The above approaches recovered transcript fragments, but not entire genes
(figure in suppl?)
To construct a more comprehensive gene set containing all of the
assembled contigs, we tried using Oases-M to merge the assemblies from
multiple $k$ values \cite{Schulz:2012je}.  While it has been
demonstrated that merged transcripts from multiple-k assemblies
contain more isoforms than those from any single $k$, the sensitivity
of Oases-M for recovering splice variants has not been fully
evaluated.
% @CTB what does this mean? ``not fully evaluated''?
We merged transcripts from our global assembly, above,
with Oases-M using a k-mer size of 27, and compared them with the
unmerged transcripts.  We then cross-validated using publicly
available ESTs, which were not used in our assembly.
The results show that
Oases-M and the unmerged assembly share about 95\% of the predicted,
splice junctions, with 5\% disjoint.  Of these 5\%,
approximately 6\% of the Oases-M-specific splice junctions are
independently supported by ESTs, while 19\% of the unmerged splice
junctions are supported by ESTs.
% @CTB change from percent to total number, with percent in parantheses
This suggests that Oases-M probably
discards a number of real splice variants, although the unmerged
assembly is also missing some found by Oases-M.

@@

We developed an exon graph approach to construct gene models from
alignments of transcripts against the genome.  Our approach,
implemented in a software package called Gimme, merges transcripts and
gene models based on overlapping exons using an exon-graph approach
(see Materials and Methods; Figure XX).  We used Gimme to obtain
20,439 gene models containing 25,859 isoforms from our global
assembly; 21,963 gene models containing 25,859 from our local
assembly; and 20,922 gene models containing 27,732 isoforms from the
merged global and local assemblies (Table~\ref{genes_transcripts}).

% (more here - what? @CTB)

@@

Our pipeline predicts many gene models and isoforms after assembly.
We validated these gene models in several different ways.  XX more!

@@

We used bowtie to map single-end reads from the source datasets to the
transcripts.  More than 80\% of the original reads could be aligned to
the transcripts, demonstrating that we did not lose a significant
amount of information during the merge process
(Table~\ref{single-end_map}).  More importantly, we also mapped
paired-end reads from technical replicates to the same gene models,
and found that more than 85\% of the paired-ends mapped concordantly
the gene models (Table~\ref{paired-end_map}).  Most of the reads that
did not map were either highly erroneous or contained low-complexity
artifactual sequence that probably originated from sample processing
and reverse transcription (cite).
% @CTB do we have something we can cite for typical % mapping?  Maybe Trinity
% paper?  How about a citation that suggests that there's a lot of crap
% in mRNAseq?
Thus the merged gene models produced by Gimme represent the significant
majority of the assemblable data.

@@

To validate the splice junctions reconstructed by the Gimme pipeline,
we used Bowtie to map mRNAseq reads directly to the transcript sequences
derived from gene models \cite{Langmead:2009fv}.  Because the
Velvet/Oases assembly pipeline does not make use of the reference
genome, reads that map across a splice junction constitute independent
verification of a splice junction's presence in a transcript.

Of 104,827 splice junctions from the gene models, 248 (0.23\%)
junctions have no spliced reads and only 484 (0.46\%) junctions
have fewer than 4 spliced reads (Fig.~\ref{cdf_single_splice}).
% @CTB how do you get junctions with no spliced reads!?
More than 99\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
real splice junctions.
% @CTB I'm slightly bugged by this statement.  Are we right?

@@

Of the 104,827 splice junctions in our gene models, 82,908 (80.31\%)
are supported by ESTs or mRNAs from Genbank.  This is especially
surprising since our mRNAseq data is from spleen, and most of the
publicly available ESTs or mRNAs are from other tissues.  Note that this
cross-validation suggests that the 19,981 novel splice junctions {\em not}
seen in publicly available ESTs and mRNAs are also likely to be real
splice junctions from spleen.
% @CTB supported by ESTs OR mRNAs, or ESTs AND mRNAs?  Surely the former?

@@

The ENSEMBL pipeline does not, however, include a large number of
splice junctions from ESTs (99,755) or mRNAs (13,641).  Cufflinks and
Gimme each recover about 10\% of these, with more than 2/3 of these
recovered by both Cufflinks and Gimme.  This indicates that both Gimme
and Cufflinks are equally adept at recovering novel splice junctions.
% @CTB note this suggests that Cufflinks improved recovery for ENSEMBL
% gene models is false...
% @CTB what fraction of Cufflinks splice junctions are high coverage (4+)?

% @CTB somewhere we might need to define terms: gene/locus,
% transcript, and splice junction...

@@

As shown above, Cufflinks and the assembly method detect a number of
distinct but equally valid splice junctions, which suggests that we
could obtain greater sensitivity by merging both sets of predictions.
We therefore used Gimme to merge the Cufflinks and assembly gene
models Table.~\ref{genes_transcripts}.  This resulted in a decreased
number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{cufflinks_asm}). The merged gene models recover 51\% and
66\% of splice junctions from ESTs and ENSENBL respectively, which is
10-15\% greater than that from corresponding unmerged gene models.
% @CTB what about splice junctions etc??

@@

To validate our predicted isoforms, we extracted coding sequences from
our gene models by using ESTScan\cite{Iseli:1999vd}.  ESTScan
successfully translated 22,488 of 27,732 (81.1\%) of our isoforms to
protein sequences with 50 or more amino acids.  We then searched for
homologous sequences in mouse ENSEMBL, and found that 17,620 (78.4\%)
of our isoforms from 12,399 distinct genes match mouse proteins at a
bit score of XXX.
Most of these matches have a bit-score/length ratio greater than 1,
which indicates a good agreement between chicken and mouse proteins.
% @CTB why so few match, really?

% @CTB for discussion: what's surprising about all of this? so many
% splice variants, etc?  Note we need to go over pickrell paper carefully.

@@

Most transcripts include a large part of untranslated regions (UTRs), especially $3'$ UTR in our datasets.
These regions are challenging to predict correctly solely from computational methods due to low degree of conservation [?].
RNA-Seq-derived gene models are useful for studying variations within UTR regions, which may be involved in regulation of isoform expression\cite{} (Fig.~\ref{long_utr}).

@@

The number of genes from our pipeline might be overestimated due to
fragmented transcripts.  In our datasets, read coverage is lower at
the $5'$ end due to the method used to convert mRNA to cDNA in library
preparation.  A bias of read coverage toward $3'$ end and low
expression level result in fragmented transcripts as shown in
Fig.~\ref{fragmented_transcripts}.  Moreover, spurious splice
junctions due to duplication in genome sequence can lead to a large
number of dubious isoforms.  This seems to occur in a few genome
sequences such as chromosome E64\_random.
Figure~\ref{isoforms_boxplot} shows that a majority of genes from our
gene models have only a few isoforms.

@@

\subsubsection{Cufflinks}

@@

\subsubsection*{isoforms and mouse}

@@

\section*{Tables}
\begin{table}[!ht]
\caption{
\bf{Unique sequences from global and local assembly}}
% @CTB do we want to make this base pairs instead of #s? Same for table 2.
\begin{tabular}{ccccc}
\hline
Dataset & \multicolumn{2}{c}{Total Sequence} & \multicolumn{2}{c}{Unique Sequence}\\
 & Global & Local & Global & Local\\
\hline
Line 6 uninfected & 338,353 & 335,180 & 10,068 (2.97\%) & 1,258 (0.37\%)\\
Line 6 infected & 353,485 & 319,920 & 11,639 (3.29\%)& 2,270 (0.70\%)\\
Line 7 uninfected & 335,876 & 302,443 & 10,649 (3.17\%) & 1,570 (0.51\%)\\
Line 7 infected & 371,404 & 327,491 & 11,859 (3.19?\%)& 1,199 (0.36\%)\\
\hline
%table information
\end{tabular}
%\begin{flushleft}Table Caption
%\end{flushleft}
\label{unique_sequences}

@@

\caption{
\bf{Number of total and unique splice junctions}}
\begin{tabular}{cccccc}
\hline
Method& Total & Unique & Unique/supported by ESTs \\ 
\hline
Oases-M assembly & 111,237 & 6,860 & 421 (6.1\%) \\
Unmerged assembly & 112,708 & 8,295 & 1,607 (19.4\%) \\
\hline
%table information
\end{tabular}
%\begin{flushleft}\footnotesize \textit{*Minumum set}
%\end{flushleft}
\label{Oases-M}
